---
title: 'Project 3: Analyses of daily COVID-19 cases across nations'
author: "Jiayi Shen (js5354), Siquan Wang (sw3442), Jack Yan (xy2395)"
date: "5/1/2020"
output:
  pdf_document:
    latex_engine: xelatex
    
---

# 1. Introduction

The pandemic of COVID-19 is the biggest challenge that the world is facing right now. Our lifes are all deeply affected by this public health crisis. By building a model on the growth of COVID-19 cases, we can have a better understanding of the current status and then plan future responses. Thus analyzing existing data and predicting future trajectories has become the most important task faced by public health expertises and policy makers.  
  
The objective of this project includings:  
- Develop an optmization algorithm to fit a logisitc curve to each region, using the global COVID-19 data;  
- Evaluate how appropiate it is to use such model;   
- Apply clustering methods to the logistic grwoth model parameters, and observe the patterns.  


## 2. Data
The datasets used in this project are adapted from https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports. We used the latest data updated on 04/29/2020, so that we have more data points. The data recorded the following variables:  

**Id:** Record ID

**Province/State:** The lcoal state/province of the record; 54% records do not have this info;

**Country/Region:** The country/regionoof the record;

**Lat:** Lattudiute of the record;

**Long:** Longitude of the record;

**Date:** Date of the record; from Jan 21 to March 23; 

**ConfirmedCases:** The number of confirme case on that day;

**Fatalities:** The number of death on that day;

For the purpose of fitting logistic growth curve, the main variables of interest are `Country/Region` and `ConfirmedCases`. We groupped the dataset by `Country/Region`, and pulled out `days since first case` and `cumulative confirmed cases on each day`.  

# 3. Method
## 3.1 Logisitic curves
Logisitic curves could be one way to model the trajectory of cumulative cases;  It is a parametric function with the form 
$$f(t) = \frac{a}{ 1+ exp\{-b(t-c)\}},$$
where $t$ is the days since the first infection; $a$ is the upper bound, i.e. the maxium number of cases a region can reach, $b$ is growth rate, and $c$ is the mid-point, where the curve changes from convex to concave; Each curve is uniquely defined by $(a,b,c)$. By design a logistic curve increases exponentially at begining and slows down at the end.  
  
  Consider the most commonly used loss function: Mean Squared Error (MSE), which takes the form 
$$
l = \frac{1}{n} \sum_{i=1}^n (Y_i - \hat{Y_i})^2  = \frac{1}{n} \sum_{i=1}^n (Y_i - \frac{a}{ 1+ \exp\{-b(T_i-c)\}})^2
$$
Then our goal is to minimize the above loss function with respect to $(a,b,c)$. To do this, we can apply Newton-Raphson algorithm. Let $\boldsymbol{\theta} = (a,b,c)$. Newton's method suggests to update $\boldsymbol{\theta}$ iteratively, such that the $i$th step is given by 
$$
\boldsymbol{\theta}_i = \boldsymbol{\theta}_{i-1} -[\nabla^2 l(\boldsymbol{\theta}_{i-1})]^{-1} \nabla l(\boldsymbol{\theta}_{i-1})
$$
where $\nabla l(\boldsymbol{\theta}_{i-1})$ is the gradient, and $[\nabla^2 l(\boldsymbol{\theta}_{i-1})]^{-1}$ is the Hessian matrix.  
  
  In this particular case, we replace the Hessian matrix with an identity matrix to simplify the computation and increase the efficiency. Step-halfing is also incorporated to control the step size. Then the $i$th step of our Newton algorithm is given by 

$$
\boldsymbol{\theta}_i = \boldsymbol{\theta}_{i-1} + \lambda \boldsymbol H_{i-1, p \times p} \nabla l(\boldsymbol{\theta}_{i-1})
$$
where $\boldsymbol H_{i-1, p \times p} = I_{p \times p}$, $\lambda \in (0,1)$.  
  
  The gradient vector $\nabla l(\boldsymbol{\theta}_{i-1})$ of our loss function is: 
\begin{equation}
\begin{pmatrix}
  {\partial l}/{\partial a}\\
  {\partial l}/{\partial b}\\
  {\partial l}/{\partial c}
\end{pmatrix}
=
\begin{pmatrix}
  \frac{2}{n}\sum_{i=1}^n (Y_i - \frac{a}{ 1+ \exp \{-b(T_i-c)\}}) \cdot \frac{-1}{1+\exp \{-b(T_i-c)\} }\\
  \frac{2}{n}\sum_{i=1}^n (Y_i - \frac{a}{ 1+ \exp \{-b(T_i-c)\}}) \cdot \frac{-a(T_i-c) \exp\{-b(T_i-c)\})}{(1+\exp \{-b(T_i-c)\})^2} \\
  \frac{2}{n}\sum_{i=1}^n (Y_i - \frac{a}{ 1+ \exp \{-b(T_i-c)\}}) \cdot \frac{ab \exp\{-b(T_i-c)\})}{(1+\exp \{-b(T_i-c)\})^2}
\end{pmatrix}
\end{equation}
  
We set the convergence criteria to be that the difference in MSE of two consecutive iterations is smaller than $10^{-5}$, and the maximum iteration number to be 1000. 

## 3.2 Clustering
To understand which countries/regions are similar in terms of the trajectory of COVID-19 cases, we applied two clustering methods, K-mean and Guassian mixture model, to group the fitted parameters $(\hat{a},\hat{b},\hat{c})$.  
   
  The guassian mixture model assumes that $\{\mathbf x_1,\mathbf x_2,...,\mathbf x_n \} \in \mathbb R^p$ are i.i.d. random vectors following a mixture mulitvariate normal distributions with $k$ hidden groups. In this case, $\mathbf x_i = (Y_i, T_i)'$ and $(a,b,c) \in \mathbb R^3$. And  
  
$$\mathbf x_i\sim
\begin{cases}
N(\boldsymbol \mu_1, \Sigma_1), \mbox{with probability }p_1 \\
N(\boldsymbol \mu_2, \Sigma_2), \mbox{with probability }p_2\\
\quad\quad\vdots\quad\quad,\quad\quad \vdots\\
N(\boldsymbol \mu_k, \Sigma_k), \mbox{with probability }p_k\\
\end{cases}
$$

$$\mathbf x_i\sim
\begin{cases}
N(\boldsymbol \mu_1, \Sigma_1), \mbox{with probability }p_1 \\
N(\boldsymbol \mu_2, \Sigma_2), \mbox{with probability }p_2\\
\quad\quad\vdots\quad\quad,\quad\quad \vdots\\
N(\boldsymbol \mu_k, \Sigma_k), \mbox{with probability }p_k\\
\end{cases}
$$

The density of a multivariate normal $\mathbf x_i$ is 
$$f(\mathbf x; \boldsymbol \mu, \Sigma) = \frac{\text{exp} (-\frac{1}{2}(\boldsymbol x - \boldsymbol \mu)^T\Sigma^{-1}(\boldsymbol x - \boldsymbol \mu))}{\sqrt{{(2\pi)^p|\Sigma|}}}$$
The observed likelihood of $\{\mathbf x_1,\mathbf x_2,...,\mathbf x_n \}$ is 

$$L(\theta; \mathbf x_1,\mathbf x_2,...,\mathbf x_n) = \prod_{i=1}^n \sum_{j = 1}^k p_j f(\mathbf x_i; \boldsymbol \mu_j, \Sigma_j)$$
Let $\mathbf r_i = (r_{i,1},...,r_{i,k})\in \mathbb R^k$  as the cluster indicator of $\mathbf x_i$, which takes form $(0, 0,...,0,1,0,0)$ with $r_{i,j} =I\{ \mathbf x_i\mbox{ belongs to  cluster } j\}$. The cluster indicator $\mathbf r_i$ is a latent variable that cannot be observed. Therefore, we use EM algorithm to iteratively estimate model parameters.  

**E-step**: Evaluate the responsibilities using the current parameter values

$$\gamma_{i, k} ^{(t)}= P(r_{i,k}=1 |\mathbf x_i,  \theta^{(t)}) =  
\frac{p_k^{(t)}f(\mathbf x_i|\boldsymbol \mu_k^{(t)}, \Sigma_k^{(t)})}
{\sum_{j=1}^K p_k^{(t)} f(\mathbf x_i|\boldsymbol \mu_j^{(t)}, \Sigma_j^{(t)})}$$

**M-step**:   
$\theta^{(t+1)} = \arg\max\ell( \mathbf{x}, \mathbf{\gamma}^{(t)}, \theta )$.  
Let $n_k = \sum_{i=1}^n \gamma_{i, k}$, we have

$$\boldsymbol \mu_k^{(t+1)} = \frac{1}{n_k} \sum_{i=1}^n \gamma_{i, k} \mathbf x_i$$

$$\Sigma_k^{(t+1)} = \frac{1}{n_k} \sum_{i=1}^n \gamma_{i, k} (\mathbf x_i - \boldsymbol \mu_k^{(t+1)})(\mathbf x_i - \boldsymbol \mu_k^{(t+1)})^T$$

$$p_k^{(t+1)} = \frac{n_k}{n}$$

By iteratively updating the E-step and the M-step, we can reach a solution upon convergence.  
  
On the other hand, the $K$-means algorithm essentially finds cluster centers and cluster assignments that minimize the objective function
$$J(\mathbf r, \boldsymbol \mu) = \sum_{i=1}^n\sum_{j=1}^kr_{i,j}\|\mathbf x_i-\mu_k\|^2$$
where $\{\boldsymbol \mu_1, \boldsymbol \mu_2,...,\boldsymbol \mu_k\}$ are the centers of the $k$ (unknown) clusters, and $\mathbf r_i = (r_{i,1},...,r_{i,k})\in \mathbb R^k$  as the *hard* cluster assignment of $\mathbf x_i$.  

# 4. Results

## 4.1 Logistic curve

```{r, eval = T, include = F, message=F}
library(tidyverse)
logisticf = function(X, a, b, c) {
  estuff = exp(-b*(X-c)) 
  return(a/(1+estuff))
}
# read in data and result
covid = readRDS("covid.RDS")
res_df = readRDS("logistic_res.RDS") %>% rename(country_region = country)

# how many regions have passed the midpoint?
sum(res_df$c < 98) # 148

# how many regions are approaching the end of the virus speading
combined_df = covid %>% 
  group_by(country_region) %>% 
  filter(date == max(date)) %>% 
  left_join(., res_df, by = "country_region")

sum(combined_df$confirmed_cases >= 0.95*combined_df$a & res_df$c < 98)

# Which regions have faster growth rates and which regions have more "flat" growth? 
# combined_df %>% filter(b > 0 & b <0.1)
# combined_df %>% filter(b >0.3 & b < 0.4)
```


Logistic curves were used to fit cumulative confirmed cases in each country/region. The fitted results for some selected countries are shown in **Figure 1**. According to the model, as of 04/29/2020, `r sum(res_df$c > 98)` regions have passed the midpoint, among which 100 regions are close to the end of virus spreading. If the confirmed cases in a region is greater than 95% of the estimated `a` (upper bound) in the logistic curve, we define the region as close to the end of virus spreading. 

In general, Regions with small population size or at an early stage of spreading tend to have a more flat rate of growth, such as Bhutan and Venezuela (**Figure 2**). On the other hand, some regions that are close or have reached the end of virus spreading (at least according to the data, **Figure 3**) tend to have a more steep rate of growth.



## Task 1.2
*Do you think if the logistic curve is a reasonable model for fitting the curmulative casesa and predicting future new cases?*  
  
I think logistic curve might not be an appropriate model for fitting the curmulative cases and predicting future new cases. First our Newton's algorithm suggests that although we spent much efforts in optimizing this optimization algorithm (including standardization, replacing Hessian by Identity mattrix and step-halfing), the convergence speed might still have some unknown internal issue and the results highly depending on the initial starting value if we do not do standardization, which means that our model is not that robust and the estimated curve shape parameters might be questionable in some unusual case. Second, some African countries' results are little bit strange base on our model, which might be due to too liitle data available.  Finally, I may suggest using some other exponential-distribution based curves to fit the data and incorporate more shape paramters in the model fitting process. Or we could some hybrid approaches to fit different types of curves based on preclassification of countries. 


## Task 2: Clustering your fitted curves

  Apply K-mean and Guassian mixture model (with EM algorithm) to cluster the fitted parameters $(\hat{a},\hat{b},\hat{c})$; Which algoirhtm does a better job in clustering those curves? Are the resulting clusters related to geogrpahic regions, or the starting timeing of the local virus speading, or the resources of the regions? You may use external informations to help understand the clusters, i.e. find plausible explanations why some regions have similar $(a, b,c)$?


\newpage

# Figures
```{r, echo=FALSE, message=FALSE}
library(patchwork)
library(tidyverse)
# read in data
covid = readRDS("covid.RDS")
# read in result
res_df = readRDS("logistic_res.RDS")

# function to plot curves
plot_logistic_curve = function(place) {
  a = res_df[res_df$country == place, 2]
  b = res_df[res_df$country == place, 3]
  c = res_df[res_df$country == place, 4]
  x = 1:120
  df = tibble(x = x, y = logisticf(x, a = a, b = b, c = c)) 
  ggplot() +
    geom_line(aes(x = x, y = y), data = df, color = "blue", size = 1.2, alpha = 0.8) +
    geom_point(aes(x = days, y = confirmed_cases),  alpha = 0.5, size = 0.8, 
               filter(covid, country_region == place)) +
    ggtitle(place, subtitle= paste0("a = ", round(a,0), ", b = ", round(b,2), ", c = ", round(c,0)))
}

plot1 = plot_logistic_curve(place = "Brunei")
plot2 = plot_logistic_curve(place = "Albania")
plot3 = plot_logistic_curve(place = "Australia")
plot4 = plot_logistic_curve(place = "France")
(plot1 + plot2) / (plot3 + plot4)

# plot_logistic_curve(place = "Venezuela")

# saveRDS(res_df, "logistic_res.RDS")

#i = i+1
#plot_logistic_curve(place = place_list[i])
```

\center __Figure 1__ Logistic curves fitted on cumulative confirmed cases in selected countries \center

\newpage


```{r echo=FALSE, message=FALSE}
plot1 = plot_logistic_curve(place = "Bhutan")
plot2 = plot_logistic_curve(place = "Venezuela")
plot3 = plot_logistic_curve(place = "Mongolia")
plot4 = plot_logistic_curve(place = "Albania")
(plot1 + plot2) / (plot3 + plot4)
```
\center __Figure 2__ Logistic curves of selected countries with flat rate of growth \center
\newpage


```{r echo=FALSE, message=FALSE}
plot1 = plot_logistic_curve(place = "Cambodia")
plot2 = plot_logistic_curve(place = "Saint Lucia")
plot1 + plot2
```
\center __Figure 3__ Logistic curves of selected countries with steep rate of growth \center
\newpage
